\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{ragged2e}

\newcolumntype{P}[1]{>{\RaggedRight\arraybackslash}p{#1}}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  citecolor=blue
}

\title{Dynamic Heuristic Shaping Improves Qualitative Research Synthesis}
\author{Henry Carstens}
\date{June 16, 2026}

\begin{document}
\maketitle

\begin{abstract}
This experiment evaluates whether heuristic shaping improves the quality of
research-agent syntheses. Eleven research-agent configurations were run under
three conditions: a deterministic control, deterministic heuristic shaping, and
heuristic shaping with LLM-assisted extraction. Outputs were judged blindly by a
multi-model LLM panel using a fixed research-quality rubric. Heuristic shaping
beat the control condition on 10 of 11 questions. The LLM-assisted heuristic
condition ranked first on 8 of 11 questions and achieved a global Elo rating of
1725.3, compared with 1541.5 for deterministic heuristic shaping and 1233.2 for
control. The result supports using heuristic shaping as the default synthesis
path for qualitative research tasks, while preserving separate forecast
calibration tests for claims about predictive accuracy.
\end{abstract}

\section{Research Question}

The experiment asks whether Phase F heuristic shaping, with and without F7
LLM-assisted extraction, produces better qualitative research syntheses than the
deterministic control path.

The evaluated configs are landscape and architecture syntheses rather than
resolved binary forecasts. Brier scoring remains the preferred method for
resolved forecast backtests. This experiment instead measures judged synthesis
quality: fit to the question, evidence structure, falsifiability, and decision
usefulness.

\section{Experimental Design}

\subsection{Conditions}

Each research-agent config was run under three conditions using the same source
set, fixed clock, and schedule configuration.

\begin{center}
\begin{tabular}{lll}
\toprule
Condition & Heuristic researcher & LLM extraction \\
\midrule
\texttt{control} & \texttt{off} & disabled \\
\texttt{heuristic} & \texttt{shaping} & disabled \\
\texttt{heuristic\_llm} & \texttt{shaping} & enabled with provider fallback \\
\bottomrule
\end{tabular}
\end{center}

The LLM fallback order was xAI, Gemini, Codex, then Anthropic. The judged
artifact for every run was \texttt{domain\_writer/summary.md}.

\subsection{Suite}

The suite contained 11 research-agent schedules:

\begin{center}
\begin{tabular}{l}
\toprule
Schedule \\
\midrule
\texttt{daily\_financial\_synthetic\_fiij} \\
\texttt{eeg\_decryption\_financial\_ts} \\
\texttt{ha\_social\_medium\_frequency\_trading} \\
\texttt{kalman\_filter\_optimality} \\
\texttt{llm\_near\_frontier\_ml\_clustering} \\
\texttt{llm\_time\_series\_baseline} \\
\texttt{multiuser\_backtest\_scaffolding} \\
\texttt{social\_media\_stationarity\_stocks} \\
\texttt{social\_stock\_data\_feeds} \\
\texttt{synthetic\_data\_fiij\_football} \\
\texttt{time\_series\_forecasting\_state} \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Blind Judge Panel}

The panel was configured with xAI, Gemini, Codex, and Anthropic. For each
question, judges received anonymized summaries and returned a ranking in the
form \texttt{RANKING: A, B, C}. Two to three judges succeeded per question.

Judges ranked outputs against the rubric implemented in
\texttt{forecasting/research\_agent/judge\_panel.py}, covering question fit,
evidence structure, falsifiability, and practical usefulness.

Consensus rankings were computed using Borda count across successful judges.
Inter-judge agreement was measured with Kendall's W. Global condition strength
was estimated with sequential Elo in \texttt{forecasting/research\_agent/elo.py},
using K=32 and an initial rating of 1500. Each judge ranking and each Borda
consensus ranking was expanded into pairwise wins.

\subsection{Execution}

The full suite was run with:

\begin{verbatim}
python -m forecasting.research_agent.evaluate_qualitative --suite \
  --suite-output runs/research/qualitative/20260616
\end{verbatim}

Elo was recomputed without re-judging with:

\begin{verbatim}
python -m forecasting.research_agent.evaluate_qualitative \
  --backfill-elo runs/research/qualitative/20260616
\end{verbatim}

Artifacts are stored under \texttt{runs/research/qualitative/20260616/},
including variant runs, per-question \texttt{qualitative\_report.json},
\texttt{suite\_report.json}, and \texttt{elo\_ratings.json}.

\section{Results}

\subsection{Per-Question Outcomes}

Heuristic shaping beat the control condition on 10 of 11 questions. The only
control win occurred in \texttt{llm\_time\_series\_baseline}, where the control
path used a specialized financial time-series baseline writer.

\scriptsize
\begin{longtable}{P{0.25\linewidth}P{0.32\linewidth}rrP{0.16\linewidth}}
\toprule
Schedule & Consensus ranking & W & Panel & Verdict \\
\midrule
\endfirsthead
\toprule
Schedule & Consensus ranking & W & Panel & Verdict \\
\midrule
\endhead
\texttt{daily\_financial\_synthetic\_fiij} & heuristic\_llm > heuristic > control & 0.75 & 2 & heuristic beats control \\
\texttt{eeg\_decryption\_financial\_ts} & heuristic > heuristic\_llm > control & 0.44 & 3 & heuristic beats control \\
\texttt{ha\_social\_medium\_frequency\_trading} & heuristic\_llm > heuristic > control & 1.00 & 3 & heuristic beats control \\
\texttt{kalman\_filter\_optimality} & heuristic\_llm > heuristic > control & 0.11 & 3 & heuristic beats control \\
\texttt{llm\_near\_frontier\_ml\_clustering} & heuristic\_llm > heuristic > control & 0.75 & 2 & heuristic beats control \\
\texttt{llm\_time\_series\_baseline} & control > heuristic > heuristic\_llm & 0.44 & 3 & control wins \\
\texttt{multiuser\_backtest\_scaffolding} & heuristic\_llm > heuristic > control & 1.00 & 3 & heuristic beats control \\
\texttt{social\_media\_stationarity\_stocks} & heuristic > heuristic\_llm > control & 0.75 & 2 & heuristic beats control \\
\texttt{social\_stock\_data\_feeds} & heuristic\_llm > heuristic > control & 1.00 & 3 & heuristic beats control \\
\texttt{synthetic\_data\_fiij\_football} & heuristic\_llm > heuristic > control & 0.78 & 3 & heuristic beats control \\
\texttt{time\_series\_forecasting\_state} & heuristic\_llm > heuristic > control & 1.00 & 2 & heuristic beats control \\
\bottomrule
\end{longtable}
\normalsize

\subsection{Aggregate Elo}

The global Elo pool shows a large separation between the two heuristic
conditions and the control condition.

\begin{center}
\begin{tabular}{lrlrr}
\toprule
Condition & Elo & W-L & Matches & Delta vs control \\
\midrule
\texttt{heuristic\_llm} & \textbf{1725.3} & 62-18 & 80 & \textbf{+492.1} \\
\texttt{heuristic} & 1541.5 & 48-32 & 80 & +308.3 \\
\texttt{control} & 1233.2 & 10-70 & 80 & baseline \\
\bottomrule
\end{tabular}
\end{center}

The mean Elo across the three conditions is 1500.0 by construction.

\subsection{Summary Statistics}

\begin{center}
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
Questions judged & 11 \\
Heuristic beats control & \textbf{10 / 11} \\
\texttt{heuristic\_llm} ranked first & \textbf{8 / 11} \\
\texttt{heuristic} ranked first & 2 / 11 \\
\texttt{control} ranked first & 1 / 11 \\
Mean Kendall's W & 0.73 \\
Elo spread, \texttt{heuristic\_llm} minus \texttt{control} & +492.1 \\
\bottomrule
\end{tabular}
\end{center}

\section{Interpretation}

The results indicate that heuristic shaping improves qualitative research
synthesis in this suite. The control condition was usually ranked last, while
the two heuristic conditions captured nearly all first-place rankings. The
largest gain came from \texttt{heuristic\_llm}, suggesting that LLM-assisted
extraction adds useful coverage when the system is asked to identify claims,
counterclaims, base rates, and decision-relevant implications from a source set.

The deterministic \texttt{heuristic} condition also performed well. Its 1541.5
Elo and 48-32 pairwise record show that the benefit is not solely attributable
to adding more LLM calls. The shaping layer appears to provide useful structure
even when claim extraction is deterministic.

The \texttt{llm\_time\_series\_baseline} exception is informative. In that
schedule, the control path used a specialized financial time-series baseline
template. Generic heuristic shaping should not replace this dedicated path for
financial-markets LLM-baseline questions without further testing.

\section{Limitations}

The experiment measures judged synthesis quality, not realized forecast
accuracy. It should be treated as a qualitative research benchmark rather than a
calibration result.

The judge panel was partial. Gemini frequently failed to emit a parseable
\texttt{RANKING:} line, and Anthropic returned authentication errors on some
calls. The mean successful panel size was about 2.5 of 4 judges per question.

Judge agreement was uneven. Mean Kendall's W was 0.73, but the Kalman question
had W=0.11, and the EEG and LLM-baseline questions each had W=0.44. Conclusions
for low-agreement questions should be treated as provisional.

The suite is small and internally generated. It is useful for regression testing
the research-agent stack, but external validity requires additional domains,
larger panels, and forecast tasks with resolved outcomes.

\section{Recommendations}

\begin{enumerate}[leftmargin=*]
\item Use \texttt{heuristic\_researcher: shaping} with LLM extraction as the
default for non-financial qualitative research syntheses.
\item Preserve specialized templates where they encode domain-specific
structure, especially the financial time-series LLM baseline path.
\item Keep Brier-scored forecast backtests as the evidence tier for predictive
accuracy claims.
\item Fix judge-infrastructure failures and re-run the suite to estimate Elo
stability.
\item Inspect low-agreement questions individually before promoting their
results as strong evidence.
\end{enumerate}

\section{Reproducibility and Data Availability}

The primary generated artifacts are local run outputs under
\texttt{runs/research/qualitative/20260616/}. The durable methods and
implementation files are version-controlled in this repository. The experiment
can be rerun from the command in Section 2.4, provided the configured LLM
providers are available.

\begin{center}
\begin{tabular}{ll}
\toprule
Item & Path \\
\midrule
Suite report & \texttt{runs/research/qualitative/20260616/suite\_report.json} \\
Elo ratings & \texttt{runs/research/qualitative/20260616/elo\_ratings.json} \\
Per-question reports & \texttt{runs/research/qualitative/20260616/*/qualitative\_report.json} \\
Variant runs & \texttt{runs/research/qualitative/20260616/*/variants/} \\
Elo implementation & \texttt{forecasting/research\_agent/elo.py} \\
Judge panel & \texttt{forecasting/research\_agent/judge\_panel.py} \\
Suite runner & \texttt{forecasting/research\_agent/evaluate\_qualitative.py} \\
\bottomrule
\end{tabular}
\end{center}

\section{Citation}

Carstens, H. (2026). \textit{Dynamic heuristic shaping improves qualitative
research synthesis}. Forecasting research notes, 2026-06-16.

\end{document}
